Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-22 examples_C.tex
updated notes for tasks after discussion with Ismail

  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{entry}\},m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$(*\label{line:precondition_entry_out}*)

@@

@@

\subsubsection{Address Space Invariant: Identity Mappings and Conditional Page Table Ownership}
\label{subsec:identitymappingsC}
Assembly-level verification of compiler output from Figure \ref{fig:calltopteinitializeC} is verbose but doable
Section \ref{sec:logic}'s logic, but only after resolving two critical challenges.\todo{Ismail, not sure if this ``verbose but doable''
 is a good framing, we should discuss
}
Two key challenges stand out and end up affecting both the pre- and post-conditions, neither of which has been addressed by prior work.
First, the update to the memory at (virtual) address \lstinline|entry| depends on subtle ownership invariants:
if the entry is present then its fractional ownership is shared with a large number of \lstinline|L4_L1_PointsTo| assertions
from the address space invariant (Figure \ref{fig:peraspaceinvariant}),
but if the entry is absent the proof requires full ownership to update it. We resolve this by extending the address space invariant
to make the owned fraction of the entry's memory \emph{dependent on its own contents}.
Second, the conversion of physical addresses into a corresponding virtual address that can be used to modify the specific
physical location relies on subtle, never-before-formalized kernel invariants.
% \looseness=-1
%
% These two factors percolate to the precondition (for conditional fractional ownership) and postcondition
% (for physical-to-virtual conversions), as the caller essentially passes output from one call to \lstinline|pte_get_next_table|
% as input to a subsequent call to traverse multiple levels.
Since the key to solving these challenges is to extend the address space invariant, we
first discuss that invariant and the kernel designs it supports, before returning to the subtle details of
verifying lines \ref{line:install_new_entryC} and \ref{line:p2vC}.
The key idea is to establish extra invariants on physical addresses that are part of a page table ---
but to do so in a way that meshes with the existing invariants (in the informal sense) already preserved in most
unverified kernel designs.
Each of the above problems requires its own extension to the invariant, but we will discuss
physical-to-virtual conversion first, both because it dictates the organization of the invariant
and because when Line \ref{line:check_entry_present_jumpC}'s
conditional check is false that is all that is necessary for the proof; correctness of the conditional branch deals with \emph{both}
extensions.

@@

\end{array}
\]
\todo[inline]{Ismail, what's APVMappings?}
\vspace{-1em}
\caption{Address space invariant of Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings. }
  \label{fig:peraspaceinvariant_with_p2v_extensionC}
\vspace{-1em}
\end{figure*}

@@

\begin{figure}\footnotesize
  \begin{lstlisting}[language=C,mathescape,escapeinside={(*}{*)}]
/* @param entry: virtual address of a non-present entry to initialize with a new physical page */
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entry\_val} \ast \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
void pte_initialize(pte_t *entry) {
  pte_t *local = kalloc(); // Allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kallocC}*)
  entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);(*\label{line:kalloc_install}*)
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entry\_val}) }$
  $\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
  $\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$(*\label{line:initialize_post}*)
}
\end{lstlisting}
\todo[inline]{more confusion of entry as program variable holding vaddr
vs entry as logical variable for the paddr, nextpaddr, etc.
update, clarifying that it returns a value for which the mapping holds.
Also, do we need the conditional implication in this postcondition? it alloc'ed from
a region already in Xi.
And shouldn't we be excluding entry (the paddr) from Xi here?
}
\vspace{-2em}
\caption{Allocating a physical page }
\label{pteinitializespecC}
\vspace{-1em}
\end{figure}

>>>>

  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{entry}\mapsto\textsf{v}\},m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$(*\label{line:precondition_entry_out}*)

@@

\todo[inline,color=cyan]{yes update logical entry to entryp, maybe remove entry $\mapsto$ v from $\Xi$ in first line?
and fix 20}

@@

\subsubsection{Address Space Invariant: Identity Mappings and Conditional Page Table Ownership}
\label{subsec:identitymappingsC}
Assembly-level verification of compiler output from Figure \ref{fig:calltopteinitializeC} is verbose but doable
Section \ref{sec:logic}'s logic, but only after resolving two critical challenges.\todo{Ismail, not sure if this ``verbose but doable''
 is a good framing, we should discuss
}\todo[color=cyan]{maybe say something about it feeling like normal asm verification \emph{because} of the virtual points-to enabling
familiar-style proving except for specific manipulation of page tables.
Maybe mention that it's well-beyond the capabilities of prior work because they just never tried? (said nicer...)
}
Two key challenges stand out and end up affecting both the pre- and post-conditions, neither of which has been addressed by prior work.
First, the update to the memory at (virtual) address \lstinline|entry| depends on subtle ownership invariants:
if the entry is present then its fractional ownership is shared with a large number of \lstinline|L4_L1_PointsTo| assertions
from the address space invariant (Figure \ref{fig:peraspaceinvariant}),
but if the entry is absent the proof requires full ownership to update it. We resolve this by extending the address space invariant
to make the owned fraction of the entry's memory \emph{dependent on its own contents}.
Second, the conversion of physical addresses into a corresponding virtual address that can be used to modify the specific
physical location relies on subtle, never-before-formalized kernel invariants.
% \looseness=-1
%
% These two factors percolate to the precondition (for conditional fractional ownership) and postcondition
% (for physical-to-virtual conversions), as the caller essentially passes output from one call to \lstinline|pte_get_next_table|
% as input to a subsequent call to traverse multiple levels.
Since the key to solving these challenges is to extend the address space invariant, we
first discuss that invariant and the kernel designs it supports, before returning to the subtle details of
verifying lines \ref{line:install_new_entryC} and \ref{line:p2vC}.
The key idea is to establish extra invariants on physical addresses that are part of a page table ---
but to do so in a way that meshes with the existing invariants (in the informal sense) already preserved in most
unverified kernel designs.
Each of the above problems requires its own extension to the invariant, but we will discuss
physical-to-virtual conversion first, both because it dictates the organization of the invariant
and because when Line \ref{line:check_entry_present_jumpC}'s
conditional check is false that is all that is necessary for the proof; correctness of the conditional branch deals with \emph{both}
extensions.

@@

\end{array}
\]
\todo[inline]{Ismail, what's APVMappings?}
\todo[inline,color=cyan]{APVMappings is authoritative ownership of the ghost map for the current address space's $\Xi$ (identity mappings).
So the last line shouldn't just existentially quantify $\delta$, but also a $\delta'$ since the authoritative map for $\Xi$ should have its own ghost name
(and does in the Rocq)
}
\vspace{-1em}
\caption{Address space invariant of Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings. }
  \label{fig:peraspaceinvariant_with_p2v_extensionC}
\vspace{-1em}
\end{figure*}

@@

\begin{figure}\footnotesize
  \begin{lstlisting}[language=C,mathescape,escapeinside={(*}{*)}]
/* @param entry: virtual address of a non-present entry to initialize with a new physical page */
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entry\_val} \ast \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
void pte_initialize(pte_t *entry) {
  pte_t *local = kalloc(); // Allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kallocC}*)
  entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);(*\label{line:kalloc_install}*)
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entry\_val}) }$
  $\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
  $\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$(*\label{line:initialize_post}*)
}
\end{lstlisting}
\todo[inline]{more confusion of entry as program variable holding vaddr
vs entry as logical variable for the paddr, nextpaddr, etc.
update, clarifying that it returns a value for which the mapping holds.
Also, do we need the conditional implication in this postcondition? it alloc'ed from
a region already in Xi.
And shouldn't we be excluding entry (the paddr) from Xi here?
}
\todo[inline,color=cyan]{no, this proof really doesn't care if entry is in or out of Xi, no need to exclude, but maybe worth explaining}
\vspace{-2em}
\caption{Allocating a physical page }
\label{pteinitializespecC}
\vspace{-1em}
\end{figure}

### 2025-07-22 relwork.tex
updated notes for tasks after discussion with Ismail

Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts). While we have not proven such a result,
an analagous result {should} hold of our work: consider that the doubles for the \texttt{mov} instructions
that access memory behave just as one would expect for a VM-ignorant logic~\cite{Chlipala2013Bedrock}.
With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle be extended to account for pageable points-to assertions by adding additional 
disjunctions to an extended points-to definition; embedding ``regular'' separation logic into such a variant
is the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.

>>>>

Unlike our work, Kolanski and Klein prove very useful embedding theorems\todo{no, Colin fix per response} stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts). While we have not proven such a result,
an analagous result {should} hold of our work: consider that the doubles for the \texttt{mov} instructions
that access memory behave just as one would expect for a VM-ignorant logic~\cite{Chlipala2013Bedrock}.
With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle be extended to account for pageable points-to assertions by adding additional 
disjunctions to an extended points-to definition; embedding ``regular'' separation logic into such a variant
is the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.
